---
title: "Handling model coefficients for #TidyTuesday collegiate sports"
author: Julia Silge
date: '2022-04-09'
format: hugo
slug: college-sports
categories:
  - rstats
  - tidymodels
tags:
  - rstats
  - tidymodels
summary: "The tidymodels framework provides extension packages for specialized tasks such as Poisson regression. Learn how to fit a zero-inflated model for understanding how R package releases are related to number of vignettes."
---

```{r setup}
#| include: false
library(knitr)
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE, 
                      message = FALSE, echo = TRUE, dpi = 180,
                      fig.width = 8, fig.height = 5)
library(tidyverse)
library(silgelib)
theme_set(theme_plex())
update_geom_defaults("rect", list(fill = "midnightblue", alpha = 0.8))

library(wikipediapreview)
wp_init()

## if you don't have fancy fonts like IBM Plex installed, run
## theme_set(theme_minimal())
```

This is the latest in my series of [screencasts](https://juliasilge.com/category/tidymodels/) demonstrating how to use the [tidymodels](https://www.tidymodels.org/) packages. This screencast is less about predictive modeling and more about how to handle and generate model coefficients with tidymodels. Let's learn more about this using the [`#TidyTuesday` dataset](https://github.com/rfordatascience/tidytuesday) on collegiate sports in the US. 🏈

```{r}
#| echo: false
blogdown::shortcode("youtube", "5LKrMTlJvo0")
```

</br>

Here is the code I used in the video, for those who prefer reading instead of or in addition to video.

## Explore data

Our modeling goal is to understand what affects expenditures on [collegiate sports in the US](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-03-29/readme.md). How many different sports are there in this dataset?

```{r}
library(tidyverse)
sports_raw <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-03-29/sports.csv')

unique(sports_raw$sports)
```

Let's combine some of those sports categories:

```{r}
sports_parsed <- sports_raw %>%
  mutate(sports = case_when(
    str_detect(sports, "Swimming") ~ "Swimming and Diving",
    str_detect(sports, "Diving") ~ "Swimming and Diving",
    str_detect(sports, "Track") ~ "Track",
    TRUE ~ sports
  ))

unique(sports_parsed$sports)
```

Let's choose some variables to explore further and create a dataset with `bind_rows()` that has one row for each sport and gender.

```{r}
sports <- bind_rows(
  sports_parsed %>%
    select(year, institution_name, sports, 
           participants = partic_men, 
           revenue = rev_men, 
           expenditure = exp_men) %>%
    mutate(gender = "men"),
  sports_parsed %>% 
    select(year, institution_name, sports, 
           participants = partic_women, 
           revenue = rev_women, 
           expenditure = exp_women) %>%
    mutate(gender = "women")
) %>%
  na.omit()

sports
```

In the screencast I did more EDA, but here let's just make one exploratory plot.

```{r}
#| fig.height = 9
sports %>%
  mutate(sports = fct_reorder(sports, expenditure)) %>%
  ggplot(aes(expenditure, sports, fill = gender, color = gender)) +
  geom_boxplot(position = position_dodge(preserve = "single"), alpha = 0.2) +
  scale_x_log10(labels = scales::dollar) +
  theme(legend.position = "top") +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  labs(y = NULL, color = NULL, fill = NULL, x = "Expenditures per team")
```

Notice the log scale and those outliers for sports like football and men's basketball! 😳 It doesn't look like there is much difference between men and women for any given sport.

## Build linear models

Let's take a straightforward, "native R" approach to fitting two linear models for this data:

-   explaining expenditures based on number of participants and gender

-   the same, but adding in sport as a predictor to estimate the impact of different sports on how much money is spent per team

```{r}
ignore_sport <- 
  lm(expenditure ~ gender + participants, data = sports)
account_for_sport <- 
  lm(expenditure ~ gender + participants + sports, data = sports)
```

In tidymodels, we recommend using broom to handle the output of models like these, so we can more easily handle, manipulate, and visualize our results. Check out [Chapter 3 of *Tidy Modeling with R*](https://www.tmwr.org/base-r.html#tidiness-modeling) for more on this topic!

```{r}
library(broom)
bind_rows(
  tidy(ignore_sport) %>% mutate(sport = "ignore"), 
  tidy(account_for_sport) %>% mutate(sport = "account for sport")
) %>%
  filter(!str_detect(term, "sports"), term != "(Intercept)") %>%
  ggplot(aes(estimate, term, color = sport)) +
  geom_vline(xintercept = 0, size = 1.5, lty = 2, color = "gray50") +
  geom_errorbar(size = 1.4, alpha = 0.7,
                aes(xmin = estimate - 1.96 * std.error, xmax = estimate + 1.96 * std.error)) +
  geom_point(size = 3) +
  scale_x_continuous(labels = scales::dollar) +
  theme(legend.position="bottom") +
  scale_color_brewer(palette = "Accent") +
  labs(x = "Change in expenditures", y = NULL, color = "Include sport in model?",
       title = "Expenditures on college sports",
       subtitle = "Colleges spend less on women's sports overall\nFor the same sport, we don't see differences by gender")
```

We see here that colleges spend less per team overall on women's sports, but this isn't true when we control for sport. Basically, it's just football driving the differences between men and women! Also, when we account for sport, the increase in expenditure per participant comes down a lot.

## Bootstrap intervals

We used the standard intervals from `lm()` in the section above, but what if we're worried about the assumptions of OLS and/or just want to create more robust interval estimates? We can use [bootstrap intervals](https://moderndive.com/8-confidence-intervals.html) instead. There are [several ways to estimate bootstrap intervals](https://www.tidymodels.org/learn/statistics/bootstrap/) in tidymodels, but the [simplest is using `reg_intervals()` from rsample](https://rsample.tidymodels.org/reference/reg_intervals.html):

```{r}
library(rsample)
set.seed(123)
ignore_intervals <- 
  reg_intervals(expenditure ~ gender + participants, data = sports, times = 500)

set.seed(123)
account_for_sport_intervals <- 
  reg_intervals(expenditure ~ gender + participants + sports, data = sports, times = 500)
```

What are the estimates for the change in expenditures for each sport?

```{r}
account_for_sport_intervals %>%
  filter(str_detect(term, "sports")) %>%
  arrange(desc(.estimate))
```

The difference between football and the next sport is LARGE. Let's make a similar plot for the model coefficients as in the last section.

```{r}
bind_rows(
  ignore_intervals %>% mutate(sport = "ignore"), 
  account_for_sport_intervals %>% mutate(sport = "account for sport")
) %>%
  filter(!str_detect(term, "sports")) %>%
  ggplot(aes(.estimate, term, color = sport)) +
  geom_vline(xintercept = 0, size = 1.5, lty = 2, color = "gray50") +
  geom_errorbar(size = 1.4, alpha = 0.7,
                aes(xmin = .lower, xmax = .upper)) +
  geom_point(size = 3) +
  scale_x_continuous(labels = scales::dollar) +
  scale_color_brewer(palette = "Accent") +
  theme(legend.position="bottom") +
  labs(x = "Change in expenditures", y = NULL, color = "Include sport in model?",
       title = "Bootstrap confidence intervals for expenditures in college sports",
       subtitle = "Colleges spend less on women's sports overall\nIn any specific sport, we don't see evidence for differences")
```

This plot looks very similar, although the relative size of the intervals for gender and number of participants has changed (intervals for number of participants are larger; intervals for gender are smaller). Again, we see that overall, the expenditures per team are much less for women's sports, but that we don't have evidence for differences within individual sports.
