---
title: "Predict #TidyTuesday NYT bestsellers"
author: Julia Silge
date: '2022-05-11'
format: hugo
slug: nyt-bestsellers
categories:
  - rstats
  - tidymodels
tags:
  - rstats
  - tidymodels
summary: "Will a book be on the NYT bestseller list a long time, or a short time? We walk through how to use wordpiece tokenization for the author names, and how to deploy your model as a REST API."
---

```{r setup}
#| include: false
library(knitr)
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE, 
                      message = FALSE, echo = TRUE, dpi = 180,
                      fig.width = 8, fig.height = 5)
library(tidyverse)
library(silgelib)
theme_set(theme_plex())
update_geom_defaults("rect", list(fill = "midnightblue", alpha = 0.8))

library(wikipediapreview)
wp_init()

## if you don't have fancy fonts like IBM Plex installed, run
## theme_set(theme_minimal())
```

This is the latest in my series of [screencasts](https://juliasilge.com/category/tidymodels/) demonstrating how to use the [tidymodels](https://www.tidymodels.org/) packages.
This screencast walks through how to use wordpiece tokenization for text feature engineering, as well as how to create a REST API to deploy your model.
Let's learn more about all this using the [`#TidyTuesday` dataset](https://github.com/rfordatascience/tidytuesday) on NYT bestsellers, which comes to us via [Post45](https://doi.org/10.18737/CNJV1733p4520220211).
📚

```{r}
#| echo: false
blogdown::shortcode("youtube", "p9ndhpyBVQc")
```

</br>

Here is the code I used in the video, for those who prefer reading instead of or in addition to video.

## Explore data

Our modeling goal is to predict which [NYT bestsellers](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-05-10/readme.md) will be on the bestsellers list for a long time, based on the book's author.

```{r}
library(tidyverse)
nyt_titles <- read_tsv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-05-10/nyt_titles.tsv')

glimpse(nyt_titles)
```

How is `total_weeks` on the NYT bestseller list distributed?

```{r}
nyt_titles %>%
  ggplot(aes(total_weeks)) +
  geom_histogram(bins = 40)
```

Which authors have been on the list the *most*?

```{r}
nyt_titles %>%
  group_by(author) %>%
  summarise(
    n = n(),
    total_weeks = median(total_weeks)
  ) %>%
  arrange(-n)
```

That Danielle Steel! Amazing!!

## Build a model

Let's start our modeling by setting up our "data budget." We'll subset down to only `author` and `total_weeks`, transform the `total_weeks` variable to "long" and "short, and stratify by our outcome `total_weeks`.

```{r}
library(tidymodels)

set.seed(123)
books_split <-
  nyt_titles %>%
  transmute(
    author,
    total_weeks = if_else(total_weeks > 4, "long", "short")
  ) %>%
  na.omit() %>%
  initial_split(strata = total_weeks)
books_train <- training(books_split)
books_test <- testing(books_split)

set.seed(234)
book_folds <- vfold_cv(books_train, strata = total_weeks)
book_folds
```

How is `total_weeks` distributed?

```{r}
books_train %>% count(total_weeks)
```

Next, let's build a modeling `workflow()` with feature engineering and a linear SVM (support vector machine).
To prepare the text of the author names to be used in modeling, let's use [wordpiece tokenization](https://huggingface.co/docs/transformers/tokenizer_summary#wordpiece).
This approach to subword tokenization works by finding all the characters and then progressively learning merge rules to form subword tokens from the original characters.
When you use wordpiece, you don't find the most common subwords, but instead the subwords with the maximum likelihood.
These merge rules can be applied to new data, including new names we've never seen before.

```{r}
library(textrecipes)

svm_spec <- svm_linear(mode = "classification")

books_rec <-
  recipe(total_weeks ~ author, data = books_train) %>%
  step_tokenize_wordpiece(author, max_chars = 10) %>%
  step_tokenfilter(author, max_tokens = 100) %>%
  step_tf(author) %>%
  step_normalize(all_numeric_predictors())

## just to see how it is working:
prep(books_rec) %>% bake(new_data = NULL) %>% glimpse()

book_wf <- workflow(books_rec, svm_spec)
book_wf
```

## Evaluate, finalize, and deploy model

Now that we have our modeling workflow ready to go, let's evaluate how it performs using our resampling folds.
We need to set some custom metrics because this linear SVM does not produce class probabilities.

```{r}
doParallel::registerDoParallel()

set.seed(123)
books_metrics <- metric_set(accuracy, sens, spec)
book_rs <- fit_resamples(book_wf, resamples = book_folds, metrics = books_metrics)
collect_metrics(book_rs)
```

Not what you'd call incredibly impressive, but at least we are pretty sure there's no data leakage!
😆


Let’s use `last_fit()` to fit one final time to the **training** data and evaluate one final time on the **testing** data.

```{r}
final_rs <- last_fit(book_wf, books_split, metrics = books_metrics)
collect_metrics(final_rs)
```

Notice that this is the first time we've used the testing data. Our metrics on the testing data are about the same as from our resampling folds.

How did we do predicting the two classes?

```{r}
collect_predictions(final_rs) %>%
  conf_mat(total_weeks, .pred_class) %>%
  autoplot()
```

We are better at predicting the books that are on the list for a short time than those that are on for a long time.

If we decide this model is good to go and we want to use it in the future, we can extract out the fitted workflow. This object can be used for prediction:

```{r}
final_fitted <- extract_workflow(final_rs)
augment(final_fitted, new_data = slice_sample(books_test, n = 1))

## again:
augment(final_fitted, new_data = slice_sample(books_test, n = 1))
```

We can also examine this model (which is just linear with coefficients) to understand what drives its predictions.

```{r}
tidy(final_fitted) %>%
  slice_max(abs(estimate), n = 20) %>%
  mutate(
    term = str_remove_all(term, "tf_author_"),
    term = fct_reorder(term, abs(estimate))
  ) %>%
  ggplot(aes(x = abs(estimate), y = term, fill = estimate > 0)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_fill_discrete(labels = c("Fewer weeks", "More weeks")) +
  labs(x = "Estimate from linear SVM (absolute value)", y = NULL, 
       fill = "How many weeks on\nbestseller list?")
```

Finally, we can deploy this model as a REST API using the [vetiver](https://vetiver.tidymodels.org/) package.

```{r}
library(vetiver)
v <- vetiver_model(final_fitted, "nyt_authors")
v

library(plumber)
pr() %>%
  vetiver_api(v)
## pipe to `pr_run()` to start the API
```
