---
title: "Find high FREX and high lift words for #TidyTuesday Stranger Things dialogue"
author: Julia Silge
date: '2022-10-20'
format: hugo
slug: stranger-things
categories:
  - rstats
tags:
  - rstats
summary: "New functionality in tidytext supports identifying high FREX and high lift words from topic modeling results."
---

```{r setup}
#| include: false
library(knitr)
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE, 
                      message = FALSE, echo = TRUE, dpi = 180,
                      fig.width = 7, fig.height = 5)
library(tidyverse)
library(silgelib)
theme_set(theme_light_plex())
update_geom_defaults("rect", list(fill = "midnightblue", alpha = 0.8))
update_geom_defaults("line", list(color = "midnightblue", alpha = 0.8))

library(wikipediapreview)
wp_init()

## if you don't have fancy fonts like IBM Plex installed, run
## theme_set(theme_minimal())
```

This is the latest in my series of [screencasts](https://www.youtube.com/juliasilge)! This screencast demonstrates how to use some brand-new functionality in [tidytext](https://juliasilge.github.io/tidytext/reference/stm_tidiers.html), using this week's [`#TidyTuesday` dataset](https://github.com/rfordatascience/tidytuesday) on *Stranger Things*. 👻 

```{r}
#| echo: false
blogdown::shortcode("youtube", "2wcDYVb-2AY")
```

</br>

The code in this blog post requires the GitHub version of tidytext as of publication. Here is the code I used in the video, for those who prefer reading instead of or in addition to video.

## Explore data

Our modeling goal is to "discover" topics in [*Stranger Things*](https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-10-18) dialogue. Instead of a supervised or predictive model where our observations have labels, this is an unsupervised approach. Let's start by reading in the data, and focusing only on the show's dialogue:

```{r}
library(tidyverse)
episodes_raw <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-10-18/stranger_things_all_dialogue.csv')

dialogue <-
  episodes_raw %>%
  filter(!is.na(dialogue)) %>%
  mutate(season = paste0("season", season))

dialogue
```

To start out with, let's create a tidy, tokenized version of the dialogue.

```{r}
library(tidytext)

tidy_dialogue <-
  dialogue %>%
  unnest_tokens(word, dialogue)

tidy_dialogue
```

What words from the dialogue have the [highest log odds](https://juliasilge.github.io/tidylo/) of coming from each season?

```{r}
library(tidylo)

tidy_dialogue %>%
  count(season, word, sort = TRUE) %>%
  bind_log_odds(season, word, n) %>%
  filter(n > 20) %>%
  group_by(season) %>%
  slice_max(log_odds_weighted, n = 10) %>%
  mutate(word = reorder_within(word, log_odds_weighted, season)) %>%
  ggplot(aes(log_odds_weighted, word, fill = season)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(vars(season), scales = "free") +
  scale_y_reordered() +
  labs(y = NULL)
```

We can see that:

-   Season 1 is more about Barb 😭 and Will
-   Season 2 introduces Bob 😭😭, Dart, and the rainbow/sunflower imagery
-   Season 3 has Russians and the Scoops shop
-   Season 4 brings us Eddie, Vecna, and Yuri

Lots of proper nouns in here!

## Train a topic model

To train a topic model with the stm package, we need to create a sparse matrix from our tidy tibble of tokens. Let's treat each episode of *Stranger Things* as a document.

```{r}
dialogue_sparse <-
  tidy_dialogue %>%
  mutate(document = paste(season, episode, sep = "_")) %>%
  count(document, word) %>%
  filter(n > 5) %>%
  cast_sparse(document, word, n)

dim(dialogue_sparse)
```

This means there are 34 episodes (i.e. documents) and `r ncol(dialogue_sparse)` different tokens (i.e. terms or words) in our dataset for modeling.

A topic model like this one models:

- each **document** as a mixture of topics
- each **topic** as a mixture of words

The most important parameter when training a topic modeling is `K`, the number of topics. This is like `k` in k-means in that it is a hyperparamter of the model and we must choose this value ahead of time. We could [try multiple different values](https://juliasilge.com/blog/evaluating-stm/) to find the best value for `K`, but this is a pretty small dataset so let's just stick with `K = 5`.

```{r}
library(stm)
set.seed(123)
topic_model <- stm(dialogue_sparse, K = 5, verbose = FALSE)
```

To get a quick view of the results, we can use summary().

```{r}
summary(topic_model)
```

## Explore topic model results

To explore more deeply, we can `tidy()` the topic model results to get a dataframe that we can compute on. The `"beta"` matrix of topic-word probabilities gives us the highest probability words from each topic.

```{r}
tidy(topic_model, matrix = "beta") %>%
  group_by(topic) %>%
  slice_max(beta, n = 10, ) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  select(-beta) %>%
  pivot_wider(
    names_from = "topic", 
    names_glue = "topic {.name}",
    values_from = term
  ) %>%
  select(-rank) %>%
  knitr::kable()
```

Well, that's pretty boring, isn't it?! This can happen a lot with topic modeling; you [typically don't want to remove stop words before building topic models](http://dx.doi.org/10.1162/tacl_a_00099) but then the highest probability words look mostly the same from each topic.

People who work with topic models have come up with alternate metrics for identifying important words. One is FREX (high frequency _and_ high exclusivity) and another is lift. Look at the details at `?stm::calcfrex()` and `?stm::calclift()` to learn more about these metrics, but they measure about what they sound like they do.

Before now, there was no support in tidytext for these alternate ways of identifying important words, but I just merged in new functionality for this. To use these as of today, you will need to install from GitHub via `devtools::install_github("juliasilge/tidytext")`.

We can find high FREX words:

```{r}
tidy(topic_model, matrix = "frex") %>%
  group_by(topic) %>%
  slice_head(n = 10) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  pivot_wider(
    names_from = "topic", 
    names_glue = "topic {.name}",
    values_from = term
  ) %>%
  select(-rank) %>%
  knitr::kable()
```

Or high lift words:

```{r}
tidy(topic_model, matrix = "lift") %>%
  group_by(topic) %>%
  slice_head(n = 10) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  pivot_wider(
    names_from = "topic", 
    names_glue = "topic {.name}",
    values_from = term
  ) %>%
  select(-rank) %>%
  knitr::kable()
```

These return a ranked set of words (not the underlying metrics themselves). They give us a much clearer idea of what makes each topic unique!

To connect the topics back to seasons, let's use `tidy()` again, finding the `"gamma"` matrix of document-topic probabilities.

```{r}
episode_gamma <- tidy(
  topic_model, 
  matrix = "gamma",
  document_names = rownames(dialogue_sparse)
)
episode_gamma
```

We can parse these results to find the season info again:

```{r}
episodes_parsed <- 
  episode_gamma %>%
  separate(document, c("season", "episode"), sep = "_")

episodes_parsed
```

Let's visualize how these document-topic probabilities are distributed over the seasons.

```{r}
episodes_parsed %>%
    mutate(topic = factor(topic)) %>%
    ggplot(aes(topic, gamma, fill = topic)) +
    geom_boxplot(alpha = 0.7, show.legend = FALSE) +
    facet_wrap(vars(season)) +
    labs(y = expression(gamma))
```

Each season mostly consists of one of these topics, with season 3 consisting of more like two topics. We could also look at how topic is related to season by using `stm::estimateEffect()`, like [in this blog post](https://juliasilge.com/blog/spice-girls/).

