---
title: "Classification metrics for #TidyTuesday GPT detectors"
author: Julia Silge
date: '2023-07-19'
format: hugo
slug: gpt-detectors
categories:
  - rstats
  - tidymodels
tags:
  - rstats
  - tidymodels
summary: "Learn about different kinds of metrics for evaluating classification models, and how to compute, compare, and visualize them."
---

```{r setup}
#| include: false
library(knitr)
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE, 
                      message = FALSE, echo = TRUE, dpi = 180,
                      fig.width = 7, fig.height = 5)
library(tidyverse)
library(silgelib)
theme_set(theme_light_plex())
update_geom_defaults("rect", list(fill = "midnightblue", alpha = 0.8))
update_geom_defaults("line", list(color = "midnightblue", alpha = 0.8))

library(wikipediapreview)
wp_init()

## if you don't have fancy fonts like IBM Plex installed, run
## theme_set(theme_minimal())
```

This is the latest in my series of [screencasts](https://www.youtube.com/juliasilge)! This screencast focuses on how to use tidymodels for computing classification metrics, using this week's [`#TidyTuesday` dataset](https://github.com/rfordatascience/tidytuesday) on GPT detectors. 🤖

```{r}
#| echo: false
blogdown::shortcode("youtube", "8N5zIHSzJoE")
```

</br>

Here is the code I used in the video, for those who prefer reading instead of or in addition to video.

## Explore data

We're not going to train a model here, but instead use the output (i.e. predictions) from a handful of models that have been trained to [detect text output from GPT](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-07-18/readme.md). Let's start by reading in the data:

```{r}
library(tidyverse)
library(detectors)

detectors
```

The `kind` variable tells us whether a document was written by a human or generated via GPT, and the two `.pred_*` variables tells us what `detector` thought about that text, the predicted probability (`.pred_AI`) and predicted class (`.pred_class`) of that text being generated by AI. The `native` variable records whether a certain document was written by a native English writer or not.

```{r}
detectors |> 
  count(native, kind, .pred_class)
```

This is a great example for talking about classification metrics, because we are in a pretty common situation where a variable of interest (`native`) only applies to one class, the real humans. The AI-generated texts were certainly not generated by a native English writer, but they also were not generated by a non-native English writer.

For the real humans, what is the distribution of predicted probability for native and non-native English writers?

```{r}
#| fig-width: 7
#| fig-height: 6
detectors |> 
  filter(!is.na(native)) |> 
  mutate(native = case_when(native == "Yes" ~ "Native English writer",
                            native == "No" ~ "Non-native English writer")) |> 
  ggplot(aes(.pred_AI, fill = native)) +
  geom_histogram(bins = 40, show.legend = FALSE) +
  facet_wrap(vars(native), scales = "free_y", nrow = 2) 
```

This is the main point of the [paper this data comes from](https://arxiv.org/abs/2304.02819): GPT detectors are much more wrong for non-native English writers than for native writers. Pretty dramatic! Let's walk through how to use classification metrics to measure this in a different way from looking at the distribution overall.

These GPT detectors are classification models (predicting "human" vs. "AI"), and there are two main categories of metrics for classification models:

- Metrics that use a predicted **class**, like "human" or "AI"
- Metrics that use a predicted **probability**, like `.pred_AI = 0.828`

## Metrics that use a predicted class

In [tidymodels](https://www.tidymodels.org/), the package that handles model metrics is [yardstick](https://yardstick.tidymodels.org/). Let's start by making a confusion matrix, which uses the predicted classes, for the dataset as a whole:

```{r}
library(yardstick)

detectors |> 
  conf_mat(kind, .pred_class)
```

This $2 \times 2$ matrix or table tells us specifics about how these models are right and wrong. Overall, these models look to be better at identifying human documents (2019 right vs. 449 wrong) than identifying AI documents (1158 right vs. 2559 wrong).

One of the most common metrics for classification models is accuracy, just the plain old proportion of our data that is predicted correctly:

```{r}
detectors |> 
  accuracy(kind, .pred_class)
```

Random guessing for a binary classification problem will give you 0.5, so here we see that these detectors are not much better than random guessing, aggregated over all these documents and different models/detectors. Do some detectors do better than others?

```{r}
detectors |> 
  group_by(detector) |> 
  accuracy(kind, .pred_class)
```

There is a little bit of variation here, I guess. What about differences across the `native` variable?

```{r}
detectors |> 
  group_by(native) |> 
  accuracy(kind, .pred_class)
```

This is now a much bigger difference. Notice that the accuracy for native English writers is **GREAT** while for non-native English writers it's not much better than for the AI documents (remember that `native = NA` here means an AI generated document).

If you've learned a bit about ML, you probably have heard that accuracy is often not a great metric (especially when you have much class imbalance) so let's try out a pair of related metrics that tell us [the true positive rate and true negative rate](https://en.wikipedia.org/wiki/Sensitivity_and_specificity).

Sensitivity tells us the true positive rate:

```{r}
detectors |> 
  sensitivity(kind, .pred_class)

detectors |> 
  group_by(detector) |> 
  sensitivity(kind, .pred_class)
```

Sensitivity is a metric that can vary between 0 (bad) and 1 (good) so this looks like we will have lots of false negatives. Can we measure the true positive rate for different types of human writers?

```{r}
detectors |> 
  group_by(native) |> 
  sensitivity(kind, .pred_class)
```

No, we can't! Since there are no AI-generated documents written by native (or non-native) English writers, we can't compute any metrics that need those empty elements of the confusion matrix:

```{r}
detectors |> 
  filter(!is.na(native)) |> 
  conf_mat(kind, .pred_class)
```

Specificity, the true negative rate, works the same way, but "flipped":

```{r}
detectors |> 
  group_by(native) |> 
  specificity(kind, .pred_class)
```

We can see that the true negative rate is much better for native English writers than non-native ones, but we can't compute a true negative rate for _only_ the AI-generated documents. This kind of situation comes up a fair amount for real models where a variable you are interested in only applies to one class in your classification problem.

## Metrics that use a predicted probability

The other main category of metrics for this type of model use a probability rather than a class. If you have ever seen or used an ROC curve, it is an example of model evaluation that uses probabilities:

```{r}
detectors |> 
    roc_curve(kind, .pred_AI) |> 
    autoplot()
```

Notice how we're using `.pred_AI` instead of `.pred_class`. How do the different detectors/models do in an ROC curve?

```{r}
detectors |> 
    group_by(detector) |> 
    roc_curve(kind, .pred_AI) |> 
    autoplot()
```

Yikes, OK. These curves can be used to compute the ROC AUC (area under the ROC curve):

```{r}
detectors |> 
  group_by(detector) |> 
  roc_auc(kind, .pred_AI) 
```

Can we compute this kind of metric and compare across the `native` categories?

```{r}
#| error: true
detectors |> 
    group_by(native) |> 
    roc_curve(kind, .pred_AI)
```

Unfortunately, no, because in, say, the non-native English writer category, there are no AI-generated texts. 

Let's think about a metric that works in a different way, like [mean log loss](https://yardstick.tidymodels.org/reference/mn_log_loss.html). This metric can distinguish between predictions that are a little wrong vs. very wrong; for example, it will say that a real human document with `.pred_AI = 0.8` is predicted worse than one with `.pred_AI = 0.6`.

```{r}
detectors |> 
    mn_log_loss(kind, .pred_AI)
```

A log loss is better when it is lower. How does log loss vary over the detectors/models?

```{r}
detectors |> 
    group_by(detector) |> 
    mn_log_loss(kind, .pred_AI) |> 
    arrange(.estimate)
```

This is a pretty dramatic range in log loss. 😳 What about across the `native` variable?

```{r}
detectors |> 
    group_by(native) |> 
    mn_log_loss(kind, .pred_AI)
```

Here we see that the documents written by non-native English writers are predicted better than the AI-generated documents, but way, way worse than those by native Engish writers.
