---
title: 'Where are #TidyTuesday haunted cemeteries compared to haunted schools?'
author: Julia Silge
date: '2023-10-11'
format: hugo
slug: haunted-places
categories:
  - rstats
tags:
  - rstats
summary: "Use tidy log odds to compare which US states are more likely to have haunted cemeteries or haunted schools."
---

```{r setup}
#| include: false
library(knitr)
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE, 
                      message = FALSE, echo = TRUE, dpi = 180,
                      fig.width = 7, fig.height = 5)
library(tidyverse)
library(silgelib)
theme_set(theme_light_plex())
update_geom_defaults("rect", list(fill = "midnightblue", alpha = 0.8))
update_geom_defaults("line", list(color = "midnightblue", alpha = 0.8))

## if you don't have fancy fonts like IBM Plex installed, run
## theme_set(theme_minimal())
```

This is the latest in my series of [screencasts](https://www.youtube.com/juliasilge)! It's spooky season, and this screencast focuses on how to compute weighted log odds using this week's [`#TidyTuesday` dataset](https://github.com/rfordatascience/tidytuesday) on haunted places in the United States. 👻

```{r}
#| echo: false
blogdown::shortcode("youtube", "TfgG50P_gOs")
```

</br>

Here is the code I used in the video, for those who prefer reading instead of or in addition to video.

## Explore data

Our analysis goal here is to understand how different types of [haunted places in the United States](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-10-10/readme.md) vary by state. Let's start by reading in the data:

```{r}
library(tidyverse)
haunted_places <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-10-10/haunted_places.csv')
```

The `description` variable is longer text explaining details about the haunting, while the `location` variable is shorter text describing the haunted location. Which states are these hauntings in?

```{r}
haunted_places |> 
  count(state, sort = TRUE)
```

And what kinds of places are haunted?

```{r}
haunted_places |> 
  slice_sample(n = 10) |> 
  select(location)
```

What are the most common words used for the haunted locations?

```{r}
library(tidytext)

haunted_places |> 
  unnest_tokens(word, location) |> 
  count(word, sort = TRUE)
```

Looks like there are lots of haunted cemeteries, which seems reasonable to me, but also lots of haunted _schools_! 😱 Haunted schools??!? Won't someone think of the children? I don't approve at all.

## Weighted log odds

I'd like to know what states are more likely to have haunted cemeteries and what states are more likely to have haunted schools, so I can know which states keep their hauntings where they _belong_. Let's use [tidylo](https://juliasilge.github.io/tidylo/), a package for weighted log odds using tidy data principles. A log odds ratio is a way of expressing probabilities, and we can _weight_ a log odds ratio so that our implementation does a better job dealing with different features having different counts. Think about how the different states have different numbers of haunted places; we'd like to compute a log odds ratio that gives us a more robust estimate across the states with many haunted places and those with few. You can [read more about tidylo](https://juliasilge.com/blog/tidylo-cran/) in a previous blog post.

To start, let's create a dataset of counts:

```{r}
haunted_counts <-
  haunted_places |> 
  mutate(location = str_to_lower(location)) |> 
  mutate(location = case_when(
    str_detect(location, "cemetery|graveyard") ~ "cemetery",
    str_detect(location, "school") ~ "school",
    .default = "other"
  )) |> 
  count(state_abbrev, location)

haunted_counts
```

Now we can compute the log odds, weighted (the tidylo default) via empirical Bayes:

```{r}
#| fig-width: 10
#| fig-height: 4.5
library(tidylo)

haunted_counts |> 
  bind_log_odds(location, state_abbrev, n) |> 
  group_by(location) |> 
  slice_max(log_odds_weighted, n = 5) |> 
  mutate(state_abbrev = reorder(state_abbrev, log_odds_weighted)) |> 
  ggplot(aes(log_odds_weighted, state_abbrev, fill = location)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(vars(location), scales = "free") +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Weighted log odds (empirical Bayes)", y = NULL)
```

This plot shows us that: 

- Illinois, Idaho, and Connecticut are reasonable places where haunted locations are more likely to be cemeteries compared to other states. 
- The haunting situation in California, South Dakota, and North Dakota is _not good_ as their haunted locations are more likely to be schools. Beware! 👻
- Delaware, Washington DC, and Maryland have haunted locations that are more likely to be something other than cemeteries and schools, compared to the other states. What are some of these, specifically in DC?

```{r}
haunted_places |> 
  filter(state_abbrev == "DC") |> 
  slice_sample(n = 10) |> 
  select(location)
```

Seems like the White House is pretty haunted!
