---
title: "Getting started with k-means and #TidyTuesday employment status"
author: Julia Silge
date: '2021-02-24'
slug: kmeans-employment
categories:
  - rstats
  - tidymodels
tags:
  - rstats
  - tidymodels
subtitle: ''
summary: "Use tidy data principles to understand which kinds of occupations are most similar in terms of demographic characteristics."
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: true
projects: []
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
                      echo = TRUE, dpi = 300, cache.lazy = FALSE,
                      tidy = "styler", fig.width = 8, fig.height = 5)
library(scales)
library(tidyverse)
library(silgelib)
theme_set(theme_plex())
```


This is the latest in my series of [screencasts](https://juliasilge.com/category/tidymodels/) demonstrating how to use the [tidymodels](https://www.tidymodels.org/) packages, from starting out with first modeling steps to tuning more complex models. Today's screencast uses the [broom](https://broom.tidymodels.org/) package to tidy output from k-means clustering, with this week's [`#TidyTuesday` dataset](https://github.com/rfordatascience/tidytuesday) on employment and demographics.


```{r, echo=FALSE}
blogdown::shortcode("youtube", "opHDQzhO5Fw")
```

</br>

Here is the code I used in the video, for those who prefer reading instead of or in addition to video.

## Explore the data

Our modeling goal is to use k-means clustering to explore [employment by race and gender](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-02-23/readme.md). This is a good screencast for folks who are more new to k-means and want to understand how to apply it to a real-world data set.


```{r}
library(tidyverse)

employed <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-23/employed.csv')

```

Let's start by focusing on the **industry** and **occupation** combinations available in this data, and average over the years available. We aren't looking at any time trends, but instead at the demographic relationships.

```{r}
employed_tidy <- employed %>%
    filter(!is.na(employ_n)) %>%
    group_by(occupation = paste(industry, minor_occupation), race_gender) %>%
    summarise(n = mean(employ_n)) %>%
    ungroup()
```

Let's create a `dataframe` read for k-means. We need to center and scale the variables we are going to use, since they are on such different scales: the proportions of each category who are Asian, Black, or women and the total number of people in each category.

```{r}
employment_demo <- employed_tidy %>%
    filter(race_gender %in% c("Women", "Black or African American", "Asian")) %>%
    pivot_wider(names_from = race_gender, values_from = n, values_fill = 0) %>%
    janitor::clean_names() %>%
    left_join(employed_tidy %>%
                  filter(race_gender == "TOTAL") %>%
                  select(-race_gender) %>%
                  rename(total = n)) %>%
    filter(total > 1e3) %>%
    mutate(across(c(asian, black_or_african_american, women), ~ . / (total)),
           total = log(total),
           across(where(is.numeric), ~as.numeric(scale(.)))) %>%
    mutate(occupation = snakecase::to_snake_case(occupation))

employment_demo
```

## Implement k-means clustering

Now we can implement k-means clustering, starting out with three centers. What does the output look like?

```{r}
employment_clust <- kmeans(select(employment_demo, -occupation), centers = 3)
summary(employment_clust)
```

The original format of the output isn't as practical to deal with in many circumstances, so we can load the [broom](https://broom.tidymodels.org/) package (part of [tidymodels](https://www.tidymodels.org/)) and use verbs like `tidy()`. This will give us the centers of the clusters we found:


```{r}
library(broom)
tidy(employment_clust)
```

If we `augment()` the clustering results with our original data, we can plot any of the dimensions of our space, such as **total employed** vs. **proportion who are Black**. We can see here that there are really separable clusters but instead a smooth, continuous distribution from low to high along both dimensions. Switch out another dimension like `asian` to see that projection of the space.

```{r}
augment(employment_clust, employment_demo) %>%
    ggplot(aes(total, black_or_african_american, color = .cluster)) +
    geom_point()
```


## Choosing k

We used `k = 3` but how do we know that's right? There are lots of complicated or "more art than science" ways of choosing `k`. One way is to look at the total within-cluster sum of squares and see if it stops dropping off so quickly at some value for `k`. We can get that from another verb from broom, `glance()`; let's try lots of values for `k` and see what happens to the total sum of squares.

```{r, fig.width=7}
kclusts <- 
  tibble(k = 1:9) %>%
  mutate(
    kclust = map(k, ~kmeans(select(employment_demo, -occupation), .x)),
    glanced = map(kclust, glance),
  )

kclusts %>%
  unnest(cols = c(glanced)) %>%
  ggplot(aes(k, tot.withinss)) +
  geom_line(alpha = 0.5, size = 1.2, color = "midnightblue") +
  geom_point(size = 2, color = "midnightblue")

```

I don't see a major "elbow" `r emo::ji("muscle")` but I'd say that `k = 5` looks pretty reasonable. Let's fit k-means again.

```{r}
final_clust <- kmeans(select(employment_demo, -occupation), centers = 5)
```

To visualize this final result, let's use [plotly](https://plotly-r.com/) and add the `occupation` name to the hover so we can mouse around and see which occupations are more similar.

```{r out.width = "100%"}
library(plotly)
library(widgetframe)

p <- augment(final_clust, employment_demo) %>%
    ggplot(aes(total, women, color = .cluster, name = occupation)) +
    geom_point()
p <- ggplotly(p, height = 500)
frameWidget(p)
```


Remember that you can switch out the axes for `asian` or `black_or_african_american` to explore dimensions.



