---
title: 'Positron in action with #TidyTuesday orca encounters'
author: Julia Silge
date: '2024-10-14'
format: hugo
slug: orcas-positron
categories:
  - rstats
tags:
  - rstats
summary: "Get to know Positron, a new, next-generation data science IDE, using this week's Tidy Tuesday data on encounters with orcas."
---

```{r setup}
#| include: false
library(knitr)
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE, 
                      message = FALSE, echo = TRUE, dpi = 180,
                      fig.width = 7, fig.height = 5)
library(tidyverse)
library(silgelib)
theme_set(theme_light_plex())
update_geom_defaults("rect", list(fill = "midnightblue", alpha = 0.8))
update_geom_defaults("line", list(color = "midnightblue", alpha = 0.8))
update_geom_defaults("point", list(color = "midnightblue", alpha = 0.8))

## if you don't have fancy fonts like IBM Plex installed, run
## theme_set(theme_minimal())
```

This is my first [screencast](https://www.youtube.com/juliasilge) in quite a while! I've been busy working on [Positron](https://github.com/posit-dev/positron) a new, next-generation data science IDE, and I'm excited to share a little first look at it with you today. Positron is currently in public beta and there is still a lot of work to get it ready for everyone to use, but in this screencast, I give you a first look at what data analysis with Positron will look like, using this week's [`#TidyTuesday` dataset](https://github.com/rfordatascience/tidytuesday) on orca encounters in the Salish Sea.

```{r}
#| echo: false
blogdown::shortcode("youtube", "5BojM5EciPs")
```

Huge thanks to Jadey Ryan for her [work on this package](https://github.com/jadeynryan/orcas)!

We know Positron still has some rough edges, but if you are interested in experimenting with it, I invite you to [download the most recent beta release](https://github.com/posit-dev/positron/releases) and [share your feedback with us](https://github.com/posit-dev/positron/discussions).
